Recalibrating Gravitational Wave Phenomenological Waveform Model

### 2023-05-01 src/tex/ms.tex

We use the constant PSD weighted loss function, $\mathcal{L}_{\mathrm{mean}}$, to evaluate the
mismatch of all testing waveforms, and present the resulting distribution in
Fig. \ref{fig:q148}. The peak of the distribution has shifted towards a lower
mismatch, with a decrease of almost one order of magnitude and a 50\% reduction
in the median. When using $\mathcal{L}_{\mathrm{fl}}$, we observe a similar
improvement with a 22.9\% decrease in the median. However, the distribution lacks
a clear peak due to the problem discussed in Section \ref{subsec:optimization}.
Despite this, both distributions exhibit significant improvement, and the
optimization scheme remains unaffected.

@@

Given that the waveform model's ansatz may not be entirely compatible with NR,
and the optimization procedure is carried out over a distribution of waveforms
with varying intrinsic parameters, it is conceivable that some trade-offs in
accuracy exist between different parts of the parameter space. If this is the
cause of the high mismatch tail's failure to improve during joint-optimization,
segmenting the parameter space into smaller subspaces should alleviate this
problem. On the other hand, if the ansatz lacks the correct parameterized form
to capture the NR waveforms' behavior as a function of the intrinsic parameters,
the results will always be biased, and we should not expect any improvement,
even if we segment the parameter space during training.

Since we know intrinsic parameters play an important role in the ansatz, we
would like to investigate how intrinsic parameters affect the recalibration
process. First, we plot the parameter space of $q$ vs. $\chi_{\mathrm{PN}}$ in
Fig. \ref{fig:ps_q148_qchi}. We can see near non-spinning waveforms demonstrate
more consistent improvement, since the ansatz are developed base on non-spinning 
waveforms. Also, the original coefficients were fitted using NR waveforms with
equal or similar spin, hence the model prefers waveforms with similar spin. On 
the other hand, spinning waveforms can either improve or worsen in terms of 
mismatch base on their intrinsic parameters. To further discuss the behavior of 
non-spinning waveforms, we plot the parameter space of $\chi_1$ vs. $\chi_2$ in
Fig.~\ref{fig:ps_q148}. Waveforms along the diagonal axis, i.e. 
$\chi_1\approx\chi_2$, show good mismatch improvements as discussed above. 
Meanwhile, the top-left ($\chi_1<0$ and $\chi_2>0$) and bottom-right ($\chi_1>0$ 
and $\chi_2<0$) regions respond to optimization differently. In the top-left region, waveforms
generally improve with along optimization. However, mismatches in the bottom-right 
region do not improve after optimization. Most waveforms even turned worse after 
optimization.

@@

\begin{figure}[t]
	\script{ps_q148_quadrants.py}
	\centering
	\includegraphics[width=\columnwidth]{figures/ps_q148_quadrants.pdf}
	\caption{Parameter space of testing waveforms. Each region is fitted independently. Colorbar
	represents log difference of mismatches before and after optimization.}
	\label{fig:ps_q148_quadrant}
\end{figure}

@@

The reduced spin approximation is a major contributor to the inaccuracies
observed in the ansatz. In IMRPhenomD, this approximation employs a single spin
parameter, $\chi_{\mathrm{PN}}$, as described in Sec. \ref{sec:method}. The
parameterization of BBH mergers using a single spin parameter results in a
degeneracy within the parameter space. Specifically, black hole events with
different spins may generate the same waveform due to identical values of
$\chi_{\mathrm{PN}}$, leading to erroneous results, particularly for highly
unequal spin events. This degeneracy produces straight lines in the parameter
space with negative slopes that depend on the mass ratio, as shown in
Fig.\ref{fig:ps_q148}. Notably, the ansatz performs better in the top left
region than the bottom right along a degeneracy line. In an attempt to address
this issue, we partitioned the parameter space into four regions, as described
in Sec. \ref{sec:result}. However, even with separate optimizations for each
regions, Fig.\ref{fig:ps_q148_quadrant} indicates that the bottom-right region
exhibits similar mismatches as before, while the top-left region's performance
has improved. This observation suggests that the ansatz is specific to certain
regions of the parameter space, with a preference for BBH events featuring
$\chi_1<0$ and $\chi_2>0$.

The division of the parameter space into four regions was a simple approach
taken for practical reasons. A more systematic approach would involve the use of
level set estimation algorithms to identify regions of interest within the
parameter space. Such an algorithm can reveal additional degeneracies or issues
that may exist within the ansatz. One possible strategy is to recalibrate
individual regions of interest to achieve better results. An alternative
approach is to select regions based on degeneracy lines. However, due to the
limited number of NR waveforms available, we were unable to implement this
approach. With more NR waveforms available in the future that cover the entire
parameter space, we can perform optimization with fewer restrictions and select
regions more systematically.

Although our study primarily focused on the IMRPhenomD model, this simple yet
versatile approach can be applied to other differentiable GW models, such as the
IMRPhenomP \citep{hannam2014simple} or IMRPhenomX \citep{pratten2020setting} 
models within the same family. By jointly optimizing a new set of coefficients,
it is expected that both models can be enhanced since they share similar
construction principles to the IMRPhenomD model. For instance, they also use PN
approximants as part of the ansatz in the inspiral segment. It will be interesting 
to recalibratethe IMRPhenomX model \citep{pratten2020setting}. Because it is
parameterized by an additional anti-symmetric spin parameter, it is expected not
to exhibit the degeneracy previously described. A more detailed investigation
may provide valuable insights into the systematics of the Phenom models.
Furthermore, this approach is applicable to other GW model families, such as NR
surrogate models or EOB models introduced previously. Such an approach could 
simplify NR waveform calibration procedures and lead to the improvement of 
existing models.

>>>>

We use the constant PSD weighted loss function, $\mathcal{L}_{\mathrm{mean}}$, to evaluate the
mismatch of all testing waveforms, and present the resulting distribution in
Fig. \ref{fig:q148}. The peak of the distribution has shifted towards a lower
mismatch, with a decrease of almost one order of magnitude and a 50\% reduction
in the median. When using $\mathcal{L}_{\mathrm{fl}}$, we observe a similar

improvement with a 22.9\% decrease in the median. However, the distribution lacks
a clear peak due to the problem discussed in Section \ref{subsec:optimization}.
Despite this, both distributions exhibit significant improvement, and the
optimization scheme remains unaffected.

@@

Given that the waveform model's ansatz may not be entirely compatible with NR,
and the optimization procedure is carried out over a distribution of waveforms
with varying intrinsic parameters, it is conceivable that some trade-offs in

accuracy exist between different parts of the parameter space. If this is the
cause of the high mismatch tail's failure to improve during joint-optimization,
segmenting the parameter space into smaller subspaces should alleviate this
problem. On the other hand, if the ansatz lacks the correct parameterized form
to capture the NR waveforms' behavior as a function of the intrinsic parameters,
the results will always be biased, and we should not expect any improvement,
even if we segment the parameter space during training.

Since we know intrinsic parameters play an important role in the ansatz, we
would like to investigate how intrinsic parameters affect the recalibration
process. First, we plot the parameter space of $q$ vs. $\chi_{\mathrm{PN}}$ in
Fig. \ref{fig:ps_q148_qchi}. We can see near non-spinning waveforms demonstrate

more consistent improvement, since the ansatz are developed base on non-spinning 
waveforms. Also, the original coefficients were fitted using NR waveforms with
equal or similar spin, hence the model prefers waveforms with similar spin. On 
the other hand, spinning waveforms can either improve or worsen in terms of 
mismatch base on their intrinsic parameters. To further discuss the behavior of 
non-spinning waveforms, we plot the parameter space of $\chi_1$ vs. $\chi_2$ in
Fig.~\ref{fig:ps_q148}. Waveforms along the diagonal axis, i.e. 
$\chi_1\approx\chi_2$, show good mismatch improvements as discussed above. 
Meanwhile, the top-left ($\chi_1<0$ and $\chi_2>0$) and bottom-right ($\chi_1>0$ 
and $\chi_2<0$) regions respond to optimization differently. In the top-left region, waveforms
generally improve with along optimization. However, mismatches in the bottom-right 
region do not improve after optimization. Most waveforms even turned worse after 
optimization.

@@

\begin{figure}[t]
	\script{ps_q148_quadrants.py}
	\centering
	\includegraphics[width=\columnwidth]{figures/ps_q148_quadrants.pdf}
	\caption{Parameter space of testing waveforms. Each region is fitted independently. Colorbar

	represents log difference of mismatches before and after optimization.}
	\label{fig:ps_q148_quadrant}
\end{figure}

@@

The reduced spin approximation is a major contributor to the inaccuracies
observed in the ansatz. In IMRPhenomD, this approximation employs a single spin
parameter, $\chi_{\mathrm{PN}}$, as described in Sec. \ref{sec:method}. The
parameterization of BBH mergers using a single spin parameter results in a
degeneracy within the parameter space. Specifically, black hole events with
different spins may generate the same waveform due to identical values of
$\chi_{\mathrm{PN}}$, leading to erroneous results, particularly for highly
unequal spin events. This degeneracy produces straight lines in the parameter
space with negative slopes that depend on the mass ratio, as shown in
Fig.\ref{fig:ps_q148}. Notably, the ansatz performs better in the top left
region than the bottom right along a degeneracy line. In an attempt to address

this issue, we partitioned the parameter space into four regions, as described
in Sec. \ref{sec:result}. However, even with separate optimizations for each
regions, Fig.\ref{fig:ps_q148_quadrant} indicates that the bottom-right region
exhibits similar mismatches as before, while the top-left region's performance

has improved. This observation suggests that the ansatz is specific to certain
regions of the parameter space, with a preference for BBH events featuring
$\chi_1<0$ and $\chi_2>0$.

The division of the parameter space into four regions was a simple approach

taken for practical reasons. A more systematic approach would involve the use of
level set estimation algorithms to identify regions of interest within the
parameter space. Such an algorithm can reveal additional degeneracies or issues
that may exist within the ansatz. One possible strategy is to recalibrate
individual regions of interest to achieve better results. An alternative
approach is to select regions based on degeneracy lines. However, due to the
limited number of NR waveforms available, we were unable to implement this
approach. With more NR waveforms available in the future that cover the entire
parameter space, we can perform optimization with fewer restrictions and select
regions more systematically.

Although our study primarily focused on the IMRPhenomD model, this simple yet
versatile approach can be applied to other differentiable GW models, such as the
IMRPhenomP \citep{hannam2014simple} or IMRPhenomX \citep{pratten2020setting} 
models within the same family. By jointly optimizing a new set of coefficients,
it is expected that both models can be enhanced since they share similar
construction principles to the IMRPhenomD model. For instance, they also use PN
approximants as part of the ansatz in the inspiral segment. It will be interesting 
to recalibratethe IMRPhenomX model \citep{pratten2020setting}. Because it is

parameterized by an additional anti-symmetric spin parameter, it is expected not
to exhibit the degeneracy previously described. A more detailed investigation
may provide valuable insights into the systematics of the Phenom models.
Furthermore, this approach is applicable to other GW model families, such as NR
surrogate models or EOB models introduced previously. Such an approach could 
simplify NR waveform calibration procedures and lead to the improvement of 
existing models.
